---
title: "SpFut.processGBIF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{processGBIF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates how to use the functions in the `SpFut.processGBIF` package in sequence.

# Overview

The package contains three functions to easily download and clean GBIF data using the `rgbif` functions. 

- `downloadGBIF`
- `cleanGBIF`
- `processGBIF`

The first two function can be used in sequence or `processGBIF` can be used as a standalone (it is a wrapper for the first two).

# download_gbif and clean_gbif

Before we begin, you will need to have your GBIF username, password, and email available. They are required inputs to the functions.

First, we'll load `SpFut.processGBIF` and save our GBIF username, password, and email.

```{r setup}
library(SpFut.processGBIF)

user <- "myname"
pwd <- "MyFancyPassword"
email <- "myname@gmail.com"
```

```{r real-login, echo=F}
user <- "rmummah"
pwd <- "M@gicPizza1"
email <- "rileymummah@gmail.com"
```


We have written functions which download and clean GBIF data separately. We test this function using Northern two-lined salamander (*Eurycea bislineata*), but other species can be selected. We recommend choosing a species with a small range or limited data to start to ensure the function runs quickly. 

Let's download the raw GBIF data and clean it as two steps.

```{r download-clean}
raw <- download_gbif(scientificName = "Eurycea bislineata",
                     user = user,
                     pwd = pwd,
                     email = email)

head(raw$dat)

raw$citation


clean <- clean_gbif(raw = raw$dat,
                    startYear = 1980)

head(clean)

colnames(clean)
```


# process_gbif

We can also download and process GBIF data in one step using the `processGBIF` function. This function also includes an option to only keep "iNat" or "museum" data depending on the user's needs.

Let's use the wrapper function to download, clean, and format GBIF data. First, we'll save only the iNaturalist data from GBIF.

```{r processGBIF-iNat}
sciname <- "Eurycea bislineata"
# This is a user-input, which is saved in the output
sp.code <- "EBIS"

# To download iNat data
iNat <- process_gbif(scientificName = sciname,
                     sp.code = sp.code,
                     keep = "iNat",
                     user = user,
                     pwd = pwd,
                     email = email)

head(iNat)
```


Now let's save the museum data from GBIF

```{r processGBIF-museum}
museum <- process_gbif(scientificName = sciname,
                       sp.code = sp.code,
                       keep = "museum",
                       user = user,
                       pwd = pwd,
                       email = email)

head(museum)
```

